---
title: "3. Discretized Simultaneous Non-negative Matrix Factrozation (`dsiNMF`)"
author:
- name: Koki Tsuyuzaki
  affiliation: Laboratory for Bioinformatics Research,
    RIKEN Center for Biosystems Dynamics Research
  email: k.t.the-answer@hotmail.co.jp
date: "`r Sys.Date()`"
bibliography: bibliography.bib
package: dcTensor
output: rmarkdown::html_vignette
vignette: |
  %\VignetteIndexEntry{3. Discretized Simultaneous Non-negative Matrix Factrozation (`dsiNMF`)}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

# Introduction

In this vignette, we consider approximating non-negative multiple matrices as a product of binary (or non-negative) low-rank matrices (a.k.a., factor matrices).

Test data is available from `toyModel`.

```{r data, echo=TRUE}
library("dcTensor")
X <- dcTensor::toyModel("dsiNMF_Easy")
```

You will see that there are some blocks in the data matrices as follows.

```{r data2, echo=TRUE, fig.height=2.7, fig.width=8}
suppressMessages(library("fields"))
layout(t(1:3))
image.plot(X[[1]], main="X1", legend.mar=8)
image.plot(X[[2]], main="X2", legend.mar=8)
image.plot(X[[3]], main="X3", legend.mar=8)
```

# Binary Simultaneous Matrix Factorization (BSMF)

Here, we consider the approximation of $K$ binary data matrices $X_{k}$ ($N \times M_{k}$) as the matrix product of $W$ ($N \times J$) and $V_{k}$ (J \times $M_{k}$):

$$
X_{k} \approx W H_{k} \ \mathrm{s.t.}\ W,H_{k} \in \{0,1\}
$$

This is the combination of binary matrix factorization (BMF [@bmf]) and simultaneous non-negative matrix decomposition (siNMF [@sinmf1; @sinmf2; @sinmf3; @amari]), which is implemented by adding binary regularization against siNMF.

For the details of arguments of dsiNMF, see `?dsiNMF`. After the calculation, various objects are returned by `dsiNMF`.

See also `siNMF` function of [nnTensor](https://cran.r-project.org/package=nnTensor) package.

## Basic Usage

In BSMF, a rank parameter $J$ ($\leq \min(N, M)$) is needed to be set in advance. Other settings such as the number of iterations (`num.iter`) or factorization algorithm (`algorithm`) are also available. For the details of arguments of dsiNMF, see `?dsiNMF`. After the calculation, various objects are returned by `dsiNMF`. BSMF is achieved by specifying the binary regularization parameter as a large value like the below:

```{r bmf, echo=TRUE}
set.seed(123456)
out_dsiNMF <- dsiNMF(X, Bin_W=1E+1, Bin_H=c(1E+1, 1E+1, 1E+1), J=3)
str(out_dsiNMF, 2)
```

The reconstruction error (`RecError`) and relative error (`RelChange`, the amount of change from the reconstruction error in the previous step) can be used to diagnose whether the calculation is converged or not.

```{r conv_bmf, echo=TRUE, fig.height=4, fig.width=8}
layout(t(1:2))
plot(log10(out_dsiNMF$RecError[2:101]), type="b", main="Reconstruction Error")
plot(log10(out_dsiNMF$RelChange[2:101]), type="b", main="Relative Change")
```

The products of $W$ and $H_{k}$s show whether the original data marices are well-recovered by `dsiNMF`.

```{r rec_bmf, echo=TRUE, fig.height=5, fig.width=8}
recX <- lapply(seq_along(X), function(x){
  out_dsiNMF$W %*% t(out_dsiNMF$H[[x]])
})
layout(rbind(1:3, 4:6))
image.plot(X[[1]], main="X1", legend.mar=8)
image.plot(X[[2]], main="X2", legend.mar=8)
image.plot(X[[3]], main="X3", legend.mar=8)
image.plot(recX[[1]], main="Reconstructed X1", legend.mar=8)
image.plot(recX[[2]], main="Reconstructed X2", legend.mar=8)
image.plot(recX[[3]], main="Reconstructed X3", legend.mar=8)
```

The histograms of $H_{k}$s show that $H_{k}$s look binary.

```{r w_h_bmf, echo=TRUE, fig.height=4, fig.width=8}
layout(rbind(1:2, 3:4))
hist(out_dsiNMF$W, main="W", breaks=100)
hist(out_dsiNMF$H[[1]], main="H1", breaks=100)
hist(out_dsiNMF$H[[2]], main="H2", breaks=100)
hist(out_dsiNMF$H[[3]], main="H3", breaks=100)
```

# Semi-Binary Simultaneous Matrix Factorization (SBSMF)

Semi-Binary Simultaneous Matrix Factorization (SBSMF) is an extension of BSMF; we can select specific factor matrix (or matrices).

To demonstrate SBSMF, next we use non-negative matrices from the `nnTensor` package.

```{r data3, echo=TRUE, fig.height=3.5, fig.width=8}
suppressMessages(library("nnTensor"))
X2 <- nnTensor::toyModel("siNMF_Easy")
layout(t(1:3))
image.plot(X2[[1]], main="X1", legend.mar=8)
image.plot(X2[[2]], main="X2", legend.mar=8)
image.plot(X2[[3]], main="X3", legend.mar=8)
```

## Basic Usage

In SBSMF, a rank parameter $J$ ($\leq \min(N, M)$) is needed to be set in advance. Other settings such as the number of iterations (`num.iter`) or factorization algorithm (`algorithm`) are also available. For the details of arguments of dsiNMF, see `?dsiNMF`. After the calculation, various objects are returned by `dsiNMF`. SBSMF is achieved by specifying the binary regularization parameter as a large value like the below:

```{r sbmf, echo=TRUE}
set.seed(123456)
out_dsiNMF2 <- dsiNMF(X2, Bin_W=1E+2, J=3)
str(out_dsiNMF2, 2)
```

`RecError` and `RelChange` can be used to diagnose whether the calculation is converged or not.

```{r conv_sbmf, echo=TRUE, fig.height=4, fig.width=8}
layout(t(1:2))
plot(log10(out_dsiNMF2$RecError[2:101]), type="b", main="Reconstruction Error")
plot(log10(out_dsiNMF2$RelChange[2:101]), type="b", main="Relative Change")
```

The products of $W$ and $H_{k}$s show whether the original data is well-recovered by `dsiNMF`.

```{r rec_sbmf, echo=TRUE, fig.height=5, fig.width=8}
recX <- lapply(seq_along(X2), function(x){
  out_dsiNMF2$W %*% t(out_dsiNMF2$H[[x]])
})
layout(rbind(1:3, 4:6))
image.plot(X2[[1]], main="X1", legend.mar=8)
image.plot(X2[[2]], main="X2", legend.mar=8)
image.plot(X2[[3]], main="X3", legend.mar=8)
image.plot(recX[[1]], main="Reconstructed X1", legend.mar=8)
image.plot(recX[[2]], main="Reconstructed X2", legend.mar=8)
image.plot(recX[[3]], main="Reconstructed X3", legend.mar=8)
```

The histograms of $H_{k}$s show that all the factor matrices $H_{k}$s look binary.

```{r w_h_sbmf, echo=TRUE, fig.height=4, fig.width=8}
layout(rbind(1:2, 3:4))
hist(out_dsiNMF2$W, breaks=100)
hist(out_dsiNMF2$H[[1]], main="H1", breaks=100)
hist(out_dsiNMF2$H[[2]], main="H2", breaks=100)
hist(out_dsiNMF2$H[[3]], main="H3", breaks=100)
```

# Session Information {.unnumbered}

```{r sessionInfo, echo=FALSE}
sessionInfo()
```

# References